Appendix 2 - Fitting proposed path models using the C-statistic

### 2016-01-26 d-sep_test.Rmd
knitr output maybe?
substantial changes
new data, RStudio files, updated code in multiple files
Edit to be compatible with table of contents.

## Steps for Performing a d-Sep Test
###### (Shipley 2000, 2003, 2004, 2009)

1. Express the hypothesized causal relationships between the variables in the form of a directed acyclic graph
<img src="figures/CH2_DAG.svg" width="600px" height="600px">
2. List each of the *k* pairs of variables in the graph that do not have an arrow between them:

@@

### Example from Shipley 2009 on how to perform the d-sep test in a generalized multilevel context 
This is to demonstrate how to assess a multilevel, repeated measures path model by applying mixed models in the tests of d-separation. I am using the supplementary material, including the data provided.

##### Path Model
`Latitude` --> `Degree-days` --> `Date of bud burst` --> `Growth` --> `Live`

@@

```{r get basis set, eval=TRUE}
library(ggm)
dag <- DAG(Live ~ Growth, Growth ~ Date, Date ~ DD, DD ~ Latitude)
isAcyclic(dag)# Logical test to check if the translated graph is acyclic, though a cyclic graph should return a warning, e.g.:
DAG(x~y, y~z, z~x)
Bu <- basiSet(dag)# Get the basis set as a list of vectors. The first two items of each vector are the names of the two non-adjacent nodes, followed by the names of the nodes in the conditioning set, which may be empty (though not in this instance).
```

@@

The first independence claim is that `Date` is independent of `Latitude` after controlling for `DD`. So, I need to obtain the *precise* null probability for this claim. Choosing the correct modeling procedure requires identifying the appropriate distribution of the dependent variable, in this case `Date`. Using some helper fucntions and the `pairs()` function in the base `graphics` package I can look at the correlations and distributions simultaneously.
```{r helper functions for scatterplot matrix, echo=FALSE}
# These were copied from the help pages of the `pairs` function with minor alterations

@@

Now I want to get the fixed-effects summaries for each model, and in particular the null probability estimates (p-values) for the pair of variables in each independence claim.
```{r nlme model & probability summaries}
# These four lines were to help me figure out how to access the right part of the summary object
#sum_fit1 <- summary(fit1)
#str(sum_fit1)
#sum_fit1$coefficients$fixed# Same as fixed.effects(fit1)
#sum_fit1$tTable# Gets the full t-test table including coefficient estimates and probabilities
summary(fit1)$tTable# lat, Date independence
```
```{r}
summary(fit2)$tTable# Growth, lat independence
```
```{r}
summary(fit3)$tTable# Growth, DD indpendence
```

@@

I would like to learn how to calculate the chi-square cumulative probability based on the degrees of freedom and *C*-statistic, a la the handy tool on this [website](http://stattrek.com/online-calculator/chi-square.aspx).

It's probably simple...

>>>>

---
output: html_document
---

## Steps for Performing a d-Sep Test (Shipley 2000, 2003, 2004, 2009)

1. Express the hypothesized causal relationships between the variables in the form of a directed acyclic graph
<img src="figures/CH2_DAG.svg" width="600px" height="600px">

2. List each of the *k* pairs of variables in the graph that do not have an arrow between them:

@@

## Example from Shipley 2009 on how to perform the d-sep test in a generalized multilevel context 
This is to demonstrate how to assess a multilevel, repeated measures path model by applying mixed models in the tests of d-separation. I am using the supplementary material, including the data provided.

### Path Model
`Latitude` --> `Degree-days` --> `Date of bud burst` --> `Growth` --> `Live`

@@

### Establish Basis Set
```{r get basis set, eval=TRUE}
library(ggm)
dag <- DAG(Live ~ Growth, Growth ~ Date, Date ~ DD, DD ~ Latitude)
isAcyclic(dag)# Logical test to check if the translated graph is acyclic, though a cyclic graph should return a warning, e.g.:
DAG(x~y, y~z, z~x)
Bu <- basiSet(dag)# Get the basis set as a list of vectors. The first two items of each vector are the names of the two non-adjacent nodes, followed by the names of the nodes in the conditioning set, which may be empty (though not in this instance).
```

@@

### Assess Independence Claims
The first independence claim is that `Date` is independent of `Latitude` after controlling for `DD`. So, I need to obtain the *precise* null probability for this claim. Choosing the correct modeling procedure requires identifying the appropriate distribution of the dependent variable, in this case `Date`. Using some helper fucntions and the `pairs()` function in the base `graphics` package I can look at the correlations and distributions simultaneously.
```{r helper functions for scatterplot matrix, echo=FALSE}
# These were copied from the help pages of the `pairs` function with minor alterations

@@

Now I want to get the fixed-effects summaries for each model, and in particular the null probability estimates (p-values) for the pair of variables in each independence claim.

### Independence Claim Model Summaries
```{r nlme model & probability summaries}
# These four lines were to help me figure out how to access the right part of the summary object
#sum_fit1 <- summary(fit1)
#str(sum_fit1)
#sum_fit1$coefficients$fixed# Same as fixed.effects(fit1)
#sum_fit1$tTable# Gets the full t-test table including coefficient estimates and probabilities
summary(fit1)$tTable# lat, Date independence
```
```{r}
summary(fit2)$tTable# Growth, lat independence
```
```{r}
summary(fit3)$tTable# Growth, DD indpendence
```

@@

I would like to learn how to calculate the chi-square cumulative probability based on the degrees of freedom and *C*-statistic, a la the handy tool on this [website](http://stattrek.com/online-calculator/chi-square.aspx). It's probably simple...
